---
title: "The Marathon Data"
subtitle: "Intro to R and Some Descriptive Analyses" 
author: "Alessio Crippa"
date: "2022-09-22"
toc: true
format:
  html:
    page-layout: full
    fontsize: .9em
    html-math-method: katex
    self-contained: true
execute:
  warning: false
  echo: true
---

## Objective

This report contains code for an introduction to R and how to perform common descriptive analyses, both uni- and multi-variate, using the marathon data, which was use in the following publication [Hyponatremia among Runners in the Boston Marathon, N Engl J Med 2005;352:1550-6](https://www.nejm.org/doi/full/10.1056/NEJMoa043901) .

**Descriptive abstract**\
Hyponatremia has emerged as an important cause of race-related death and life-threatening illness among marathon runners. We studied a cohort of marathon runners to estimate the incidence of hyponatremia and to identify the principal risk factors.

## Load R packages

The following packages are used in the tutorial. If you want to reproduce the code chunks below, be sure to install them (e.g. `install.packages("tidyverse")`).

```{r setting}
#| results: hide
#| echo: false
library(tidyverse)
library(here)
library(gt)
library(scales)
library(broom)
library(Epi)
library(ggExtra)
library(arsenal)
library(gridExtra)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

## Read, explore and manipulate data

### Read data

The `marathon` data are available in a csv format at http://alecri.github.io/downloads/data/marathon.csv .

```{r}
#| output: asis
marathon <- read.csv(here("data", "raw", "marathon_raw.csv"))
gt(head(marathon)) %>% tab_options(table.font.size = px(12))
```

Many other options exist, such as:

-   basic functions (`read.table`, `read.csv`, `read.delim`)\
-   from Excel `readxl::read_excel`\
-   from SPSS `haven::read_sav`\
-   from SAS `haven::read_sas`
-   from Stata `haven::read_dta`

A comprehensive tutorial can be found [here](https://www.datacamp.com/community/tutorials/r-data-import-tutorial).

### Explore the data

```{r}
# what is the dimension of the data  
dim(marathon)
# i.e. how many obs/rows
nrow(marathon)
# how many cols/variables?
ncol(marathon)
# what's the name of the variable
colnames(marathon)
# what is the structure of the data
glimpse(marathon)
# show the first rows
head(marathon)
# quick and dirty summary
summary(marathon)
```

### Manipulate data

Data manipulations consist of a multiplicity and combinations of steps. The `dplyr` package in `tidyverse` aids this task with a collections of *verbs*.\
I will show how to:

1.  create/modify variables with the `mutate` verb.

```{r data_management}
marathon <- mutate(
  marathon, 
  # code a factor variables
  sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
  nas135 = factor(na <= 135, labels = c("na > 135", "na ≤ 135")),
  # compute new variables
  runtime = runtime/60,
  wtdiff = postwt - prewt,
  bmi = prewt/(height^2),
  # categorize continuous variables
  bmi_cat = cut(bmi, breaks = c(15, 20, 25, 33), labels = c("< 20", "20-25", ">25")),
  wtdiff_cat = cut(wtdiff, breaks = c(-5, -2, -1, 0, 1, 2, 3, 5), 
                   include.lowest = T, right = FALSE, ordered_result = TRUE)
)
```

2.  select only variables of interest with the `select` verb.

```{r select}
marathon_sub1 <- select(marathon, sex, runtime, ends_with("wt"), bmi, na)
gt(head(marathon_sub1)) %>% tab_options(table.font.size = px(12))
```

3.  select only some observations with the `filter` verb.

```{r filter}
women_sub2 <- filter(marathon_sub1, sex == "Female", runtime >= 4, bmi > 25)
gt(women_sub2) %>% tab_options(table.font.size = px(12))
```

4.  sort data according to increasing/decreasing values of some variables with the `arrange` verb.

```{r}
arrange(women_sub2, desc(na), runtime)
```

Chaining: wrap different functions inside each other

```{r}
arrange(
  filter(
    select(
      mutate(marathon, 
             sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
             runtime = runtime/60,
             bmi = prewt/(height^2)),
      sex, runtime, ends_with("wt"), bmi, na),
    sex == "Female", runtime >= 4, bmi > 25), desc(na), runtime)
```

### The %\>% (pipe) operator

This operator allows you to pipe the output from one function to the input of another function.

```{r}
marathon %>% 
  mutate(sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
         runtime = runtime/60,
         bmi = prewt/(height^2)) %>% 
  select(sex, runtime, ends_with("wt"), bmi, na) %>%
  filter(sex == "Female", runtime >= 4, bmi > 25) %>%
  arrange(desc(na), runtime)
```

## Summary statistics and graphs

### ggplot2: grammar of graphics

```{r}
ggplot(marathon, aes(wtdiff, na, col = bmi)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ sex) +
  labs(x = "Weight change, kg", y = "Sodium concentration, mmol per liter") +
  theme_bw()
```

### Univariate descriptives

```{r}
# for a continuous variable
summary(marathon$na)
quantile(marathon$na) # percentiles (or quantiles)
c("Mean" = mean(marathon$na), "SD" = sd(marathon$na),
  "Median" = median(marathon$na), "IQR" = IQR(marathon$na),
  "Min" = min(marathon$na), "Max" = max(marathon$na),
  "25th percentile" = quantile(marathon$na, .25),
  "75th percentile" = quantile(marathon$na, .75))
summarise(marathon, mean = mean(na), sd = sd(na), median = median(na), iqr = IQR(na))
# histogram (counts)
ggplot(marathon, aes(na)) + 
  geom_histogram() +
  labs(x = "Serum sodium concentration (mmol/liter)")
# histogram (density)
ggplot(marathon, aes(na)) +
  geom_histogram(stat = "density", n = 2^5) +
  geom_line(stat = "density", col = "blue", size = 1.5) +
  labs(x = "Sodium concentration, mmol per liter")
# boxplot
ggplot(marathon, aes(na)) + 
  geom_boxplot() + 
  coord_flip() +
  labs(x = "Serum sodium concentration (mmol/liter)")
# violin plot
ggplot(marathon, aes(x = "", y = na)) + 
  geom_violin(fill = "lightblue") + 
  geom_jitter(size = 1) +
  labs(x = "", y = "Serum sodium concentration (mmol/liter)")
```

```{r}
# for a categorical variable
table(marathon$nas135)
prop.table(table(marathon$nas135))
# alternative
dtab_nas135 <- count(marathon, nas135) %>% 
  mutate(
    prop = n/sum(n),
    percentage = scales::percent(prop)
  )
dtab_nas135
# barplot (counts)
ggplot(marathon, aes(nas135)) +
  geom_bar()
# barplot (proportion)
ggplot(marathon, aes(nas135)) +
  geom_bar(aes(y = ..count../sum(..count..))) +
  labs(x = "", y = "Prevalence of hyponatremia")
```

### Bivariate descriptives (association)

```{r cont_cat}
# Continuous + categorical
by(marathon$na, marathon$sex, summary)
group_by(marathon, sex) %>% 
  summarise(n_nomiss = sum(!is.na(na)), 
            mean = mean(na), 
            median = median(na),
            sd = sd(na), 
            iqr = IQR(na), 
            min = min(na), 
            max = max(na),
            p25 = quantile(na, .25),
            p75 = quantile(na, .75))
# boxplot by levels
ggplot(marathon, aes(sex, na)) +
  geom_boxplot() + 
  labs(x = "", y = "Serum sodium concentration (mmol/liter)")
ggplot(marathon, aes(na, col = sex)) +
  geom_line(stat = "density") + 
  labs(x = "Serum sodium concentration (mmol/liter)", y = "Density")
# statistical tests
test_nasex <- t.test(na ~ sex, data = marathon, var.equal = TRUE)
test_nasex
broom::tidy(test_nasex)
# equivalent linear model
lm_nasex <- lm(na ~ sex, data = marathon)
Epi::ci.lin(lm_nasex)
wilcox.test(na ~ sex, data = marathon)
```

```{r twobytwo}
# frequencies table
tab_nabmi <- table(bmi_cat = marathon$bmi_cat, nas135 = marathon$nas135)
tab_nabmi
prop.table(tab_nabmi, margin = 1)
# customizable tables
Epi::stat.table(list(nas135, bmi_cat), 
           list(count(), percent(nas135)), 
           marathon, margin = T)
# bar plot
group_by(marathon, bmi_cat) %>%
  filter(!is.na(bmi_cat)) %>% 
  count(nas135) %>%
  mutate(perc = n/sum(n)) %>%
  ggplot(aes(x = nas135, y = perc, fill = bmi_cat)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "", y = "Frequency", fill = "Categories of BMI")
# test for independence 
chisq.test(tab_nabmi)
# measure of associations
twoby2(tab_nabmi[1:2, 2:1])
twoby2(tab_nabmi[3:2, 2:1])
```

```{r two_cont}
# correlation coefficient
with(marathon, cor(na, wtdiff,  method = "pearson", use = "complete.obs"))
# scatter plot + regression line
p_nawtdiff <- ggplot(marathon, aes(wtdiff, na)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Weight change (kg) pre/post race", y = "Serum sodium concentration (mmol/liter)")
ggExtra::ggMarginal(p_nawtdiff, type = "histogram")
```

### Table 1

```{r table1}
#| results: asis
# Table 1
tab1 <- arsenal::tableby(nas135 ~ age + sex + wtdiff + bmi_cat + runtime, data = marathon)
summary(tab1)
```

## Statistical models

### Linear regression

$$E[Y|X] = \beta_0 + \beta_1X_1 + \dots + \beta_kX_k$$

```{r}
# univariate linear regression
fit1 <- lm(na ~ sex, data = marathon)
summary(fit1)
tidy(fit1, conf.int = T)
```

```{r}
# multivariable linear regression
fit2 <- lm(na ~ wtdiff + sex, data = marathon)
ci.lin(fit2)
marathon <- mutate(marathon,
                   pred_fit2 = predict(fit2, newdata = marathon))
```

```{r}
# multivariable linear regression with interaction
fit3 <- lm(na ~ wtdiff*sex, data = marathon)
ci.lin(fit3, ctr.mat = rbind(c(0, 1, 0, 0),
                             c(0, 1, 0, 1)))
marathon <- mutate(marathon,
                   pred_fit3 = predict(fit3, newdata = marathon))
```

```{r}
# graphical comparison
gridExtra::grid.arrange(
  ggplot(marathon, aes(wtdiff, pred_fit2, col = sex)) +
    geom_line() +
    labs(x = "Weight change, kg", y = "Predicted mean na"),
    ggplot(marathon, aes(wtdiff, pred_fit3, col = sex)) +
    geom_line() +
    labs(x = "Weight change, kg", y = "Predicted mean na"),
  ncol = 2
)
```

### Logistic regression

$$\log(\textrm{odds}(Y|X)) = \beta_0 + \beta_1X_1 + \dots + \beta_kX_k$$

```{r}
# univariate logistic regression
fit4 <- glm(nas135 ~ wtdiff, data = marathon, family = binomial)
ci.exp(fit4)

marathon <- marathon %>% 
  mutate(pred_p = predict(fit4, newdata = marathon, type = "response"),
         pred_odds = predict(fit4, newdata = marathon, type = "link")) 

grid.arrange(
  ggplot(marathon, aes(wtdiff, pred_p)) +
    geom_line() +
    labs(x = "Weight change, kg", y = "Predicted probability"),
  ggplot(marathon, aes(wtdiff, pred_odds)) +
    geom_line() +
    labs(x = "Weight change, kg", y = "Predicted odds"),
  ncol = 2
)
```

```{r}
# multivariable logistic regression
fit5 <- glm(nas135 ~ wtdiff + sex, data = marathon, family = binomial)
ci.exp(fit5)
mutate(marathon, pred = predict(fit5, newdata = marathon, type = "response")) %>% 
  ggplot(aes(wtdiff, pred, col = sex)) +
  geom_line() +
  labs(x = "Weight change, kg", y = "Predicted probability")
```
